---
title: "Native R audits, coverage and practical workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native R audits, coverage and practical workflows}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

ExtremaRank runs directly in R through its C++ exact kernel. Python is not a
runtime dependency. Native certificates and observed fitted-model sensitivity
have different scopes. Search caps leave unresolved questions explicit.

```{r}
library(extremarank)
effects <- matrix(c(5, 4, 6, 5, 1, 2, 1, 2, -2, -1, -3, -2), nrow = 4,
  dimnames = list(paste0("donor", 1:4), c("A", "B", "C")))
result <- extremarank_effects(effects, k = 1, budget = 1,
  search_order = "influence", max_diagnostics = 4)
summary(result)
as.data.frame(result)
plot(result)
plot(result, type = "influence")
curve <- robustness_curve(effects, budgets = 0:2, k = 1, diagnostics = FALSE)
curve$data
plot(curve)
```

Rows in `effects` are donors, columns are features. For ordinary `extremarank`
input the orientation is features by samples. Preparation, feature filtering
and ranking ties are frozen before deletion. A lower change bound of budget+1
means stability within the budget; it does not prove that the next deletion
cardinality causes a change.

```{r}
set.seed(5006)
x <- matrix(rnorm(30 * 8), 30, dimnames = list(paste0("g", 1:30), paste0("s", 1:8)))
meta <- data.frame(sample_id = colnames(x), donor_id = colnames(x),
  group = rep(c("treated", "control"), each = 4), site = rep(1:4, 2))
preflight_extremarank(x, meta, "treated", "control", design = "welch", budget = 2)
```

Storage estimates exclude model objects and arbitrary precision integer memory.
`max_dense_bytes` guards prepared sample matrices and pseudobulk output.
DelayedMatrix inputs are read in row blocks; native audits still materialize
their smaller feature-by-biological-unit matrix. Pseudobulk aggregation does
not materialize the entire delayed cell matrix. Audit and diagnostic limits
are independent. Default diagnostics are skipped at budget zero and otherwise
capped at 100; use `diagnostics="all"` with an explicit limit when needed.

```{r, eval=requireNamespace("limma", quietly=TRUE)}
fits <- refit_extremarank(x, meta, "treated", "control", model = "limma",
  design = "welch", budget = 1, k = 5, all_features = TRUE)
summary(fits)
head(as.data.frame(fits))
zero <- refit_extremarank(x, meta, "treated", "control", model = "limma",
  design = "welch", max_refits = 0)
zero$status
```

`NOT_EVALUATED` means no deletion was run. `PARTIALLY_EVALUATED` records an
incomplete stable-looking run. `OBSERVED_CHANGED` includes actual change
witnesses, even if other scenarios failed. None certifies unseen refits.

For larger models use `store_scores="disk", score_path="scores"` or
`store_scores="none"`. `checkpoint="fits-cache"` saves each completed fit;
`resume=TRUE` validates its frozen contract and can extend `max_refits`.
`workers=2` runs two PSOCK processes and collects results in scenario order.
Processes can increase RAM; the numeric input-copy guard does not bound model
working memory. `delete_by="site"` deletes whole sites. Covariates and contrasts
are fixed even if a deletion makes the design unidentifiable.

Custom repeated-measure or multi-group models can use
`refit_extremarank_custom`. Its function must refit the supplied retained data
and return every frozen feature exactly once, with `feature_id`, `score` and
`effect`. Declare biological deletion units and control any external state.
This interface has an observed sensitivity scope.

Real microarray, single-cell, protein and lipid reproductions, including failed
fits, are recorded in the repository's `results/v04` and `results/v05` folders.
The package does not establish FDR control or biological validity.
